# Reproducible Research: Peer Assessment 1
* * *
* * *

## Loading and preprocessing the data
```{r}
if (!file.exists("activity.csv")) {
  if (!file.exists("activity.zip")) {
    stop("Please make sure the file 'activity.csv' or 'activity.zip' file is in the working directory")
  } else {
    unzip("activity.zip")
  }
}

activity <- read.csv("activity.csv", colClasses=c(NA,"Date",NA))

```
Checking the loaded data
```{r}
head(activity)
```

Checking the dimension of the data
```{r}
dim(activity)
```

Check the structure of the data
```{r}
str(activity)
```
Checking the summary of the data
```{r}
summary(activity)
```
There are `r dim(activity)[1]` records and `r dim(activity)[2]` variables. The data contain `r sum(!complete.cases(activity))` missing values.

## What is mean total number of steps taken per day?
Preparing the data to answer the question

Ignoring the missing values(Removing missing velues)

```{r}
activity_nona <- activity[complete.cases(activity),]
head(activity_nona)
```

There are `r dim(activity_nona)[1]` complete cases

Using the package plyr to format and summarize.

```{r}
##Checking if the package plyr exist and if not to install it
if(require("plyr")){
  print("plyr is loaded correctly")
} else {
  print("trying to install plyr")
  install.packages("plyr")
  if(require(plyr)){
    print("plyr installed and loaded")
  } else {
    stop("could not install plyr")
  }
}
```

Transforming the data

```{r}
total_steps_day_nona <- ddply(activity_nona, .(date), summarise,
                      total_s = sum(steps))
```
#### 1. Histogram of the total number of steps taken each day
```{r plot1}
hist(total_steps_day_nona$total_s, 
     main = "Total number of steps taken each day",
     xlab = "Total Steps", col="green")
```

#### 2. To Calculate and report the mean and median total number of steps taken per day

```{r}
summary(total_steps_day_nona)

```

The `r summary(total_steps_day_nona)[4,2]` and `r summary(total_steps_day_nona)[3,2]`

## What is the average daily activity pattern?

#### 1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}
average_steps_day_nona <- ddply(activity_nona, .(interval), 
                           summarise,
                         average_s = mean(steps))
head(average_steps_day_nona)
```

```{r plot2}
with(data=average_steps_day_nona, 
      plot(interval, average_s, type="l",
      col = "blue",
      main = "Mean Total Number of Steps Taken per Day",
      ylab = "Average Steps"))
```

#### 2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
##Finding the maximum value
mx <- max(average_steps_day_nona$average_s)

##Locating the 5-minute interval which contains the maximum number of steps
interval <- average_steps_day_nona[average_steps_day_nona$average_s==mx,1]
interval
```

The 5-minute interval which contains the maximum number of steps is `r average_steps_day_nona[average_steps_day_nona$average_s==mx,1]`

## Imputing missing values

#### 1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r}
##The total number of missing values in the dataset
sum(!complete.cases(activity))
```

#### 2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.
Using the mean for the particular 5-minute interval to fill the particular missing value

#### 3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
Using impute() from package Hmisc and ddply from package plyr


```{r}
##Loading the Hmisc package and if is not going to be found it will be get installed
if(require("Hmisc")){
  print("Hmisc is loaded correctly")
} else {
  print("trying to install Hmisc")
  install.packages("Hmisc")
  if(require(Hmisc)){
    print("Hmisc installed and loaded")
  } else {
    stop("could not install Hmisc")
  }
}
```

Imputing
```{r}
activity_imp <- ddply(activity, "interval", mutate, imputed.steps = impute(steps, mean))
#Selecting the columns of interest and ordering them using order function
activity_imp <- activity_imp[order(activity_imp$date),c(4,2,3)]
dim(activity_imp)
head(activity_imp)
```

#### 4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

Preparing the data

```{r}
imp_total_steps_day <- ddply(activity_imp, .(date), summarise,
                         imp_total_s = sum(imputed.steps))
head(imp_total_steps_day)
summary(imp_total_steps_day$imp_total_s)
```
Ploting the histogram
```{r plot3}
##Histogram of the total number of steps taken each day
hist(imp_total_steps_day$imp_total_s,
     main = "Total number of steps taken each day for Imputed Data",
     xlab = "Total Steps",
     col = "yellow")
```

Mean and median total number of steps taken per day
```{r}
summary(imp_total_steps_day)

```

The `r summary(imp_total_steps_day)[4,2]` and `r summary(imp_total_steps_day)[3,2]`

Do these values differ from the estimates from the first part of the assignment?

+ Before imputing - `r summary(total_steps_day_nona)[4,2]` and `r summary(total_steps_day_nona)[3,2]`

+ After imputing - `r summary(imp_total_steps_day)[4,2]` and `r summary(imp_total_steps_day)[3,2]`

+ The means and medians values before and after imputing are almost the same

What is the impact of imputing missing data on the estimates of the total daily number of steps?

+ The imputed data has more records

## Are there differences in activity patterns between weekdays and weekends?

#### 1. Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
```{r}
head(activity_imp)

activity_imp_day <- activity_imp

for (i in 1:dim(activity_imp_day)[1])
  if (weekdays(activity_imp_day[i,2]) %in% c("Saturday","Sunday")){
      activity_imp_day[i, "day"] = "Weekend"
    } else{
        activity_imp_day[i, "day"] = "Weekday"
      }

head(activity_imp_day)
table(activity_imp_day[,4])
```

#### 2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).


Averaging across all weekday days or weekend days
```{r}

average_steps_imp_day <- ddply(activity_imp_day, .(interval,day), 
                                summarise,
                                average_s = mean(imputed.steps))
head(average_steps_imp_day)

```

Using the package ggplot2
```{r}
##Loading the ggplot2 package and if is not going to be found it will be get installed
if(require("ggplot2")){
  print("ggplot2 is loaded correctly")
} else {
  print("trying to install ggplot2")
  install.packages("ggplot2")
  if(require(ggplot2)){
    print("ggplot2 installed and loaded")
  } else {
    stop("could not install ggplot2")
  }
}
```

Ploting

```{r plot4}

ggplot(data=average_steps_imp_day, aes(x=interval, 
                      y=average_s, 
                      group=day, 
                      colour=day)) + 
  geom_line() + 
  facet_wrap(~ day, ncol=1) +
  labs(title="The Average Number of Steps Taken") +
  labs(x="Interval",y="Average Steps")

```
